Supplement: Analysis and code for 'Testing Perceptual Accuracy in Charts using Surveys

### 2023-07-24 SDSS-2023/testing-charts-accuracy.qmd
some of these changes are just github's inability to deal with individual sentences

<!-- What is our main objective? -->

\kr{Should the abstract match -- or be close to -- our SDSS short abstract?}
\hh{I don't think it has to}

What do viewers see when we show them a data chart?
A data chart -- at its core -- maps quantitative values to graphical elements representing their relative values.
Modern data visualizations are much more than a simple, objective mapping of values to a plane; they contain contextual and design elements, and are often structured to support the viewer in understanding a particular view of a set of data or specific pattern underlying the values.
The design of a data visualization impacts a viewer's ability to achieve that understanding; a poorly designed data visualization may leave viewers struggling to understand the content or context, or make it difficult to complete accurate and useful comparison of values across groups or time points.
More broadly, the design of a data visualization can change how viewers interact with the chart.

@@

Cleveland and McGill's work -- while a foundational user study in graphical perception -- utilized a small convenience sample, consisting of only a few individuals recruited from among the authors' coworkers and their spouses.
@heerCrowdsourcingGraphicalPerception2010 reproduced Cleveland and McGill's rankings using a larger sample from a crowd sourcing platform, employing a total of XXX Amazon Mechanical Turk (mTurk) workers for the study.
Crowd sourced samples were shown by @borgoCrowdsourcingInformationVisualization2017 to be biased towards more male, younger, and relatively higher education relative to the adult U.S. population as a whole; generally, convenience and opt-in samples are not representative of the general population, a common target audience for data visualization and scientific communication work.
Further, the populations' emphasis on higher education individuals also leads to results which hold for groups of individuals who may be more likely to have prior exposure to data visualization in the context of scientific communication, or more exposure to data topics in higher education, but may not hold across other groups within the population.
<!-- Bias from crowd-sourcing? sample size?  -->

\kr{The mackinlay thing feels like maybe we are overemphasizing the ranking done by cleveland/mcgill and then mackinlay - I think we could emphasize less if we want}
\hh{my point was to emphasize that Mackinlay is a foundational paper (rightfully so), but not based on any data beyond Cleveland}

@@

Our work also centers on studying data visualization design choices and their impact on viewer behavior and accuracy of viewers' responses.
We seek to answer whether it is possible to reproduce some of their findings in the context of a survey with a large, nationally-representative set of respondents, and within that context we focus on the following research questions:

@@

We present viewers with structural variations of bar charts and ask them to answer questions comparing the size of elements within those charts.
In this work, we present the series of tests we completed and the resulting findings.
The remainder of the paper is organized as follows: first, we describe the design of the visual stimuli used in our perception tests.
We then describe the population of study respondents and obtained survey sample.
Subsequently, we share our analyses of the resulting survey responses across each of our tests, including analyses on accuracy of responses and response behavior.
Finally, we discuss implications of this work and next steps.

@@

The comparison between sets of marked pairs is designed to be a difficult task, with the difference between the values represented in the two marked pieces being a just-noticeable difference.
The **Just-Noticeable Difference** (JND) is defined as the smallest difference that will be detected 50% of the time.
Prior results from studies on bar charts and pie charts [@luModelingJustNoticeable2022] inform the differences in charts shown to our survey panelists.

@@

```{=tex}
\hh{Not sure where should put this, parking it here for right now:}
\hh{expectations - start}
```
What we call 'aligned' and 'unaligned', here, is similar to Cleveland and McGill's set of rankings, but with some modifications: both 'aligned' and 'unaligned' bars share the same axis.
Aligned tiles are additionally anchored in the same position in one dimension, i.e. the difference between their sizes can be reduced to a positional assessment.
Unaligned tiles do not share this anchor, however, the context of the other tiles in the chart provide a frame, which *should* help with an assessment of the tiles' sizes beyond a comparison of (arc) lengths or areas.

We would expect that comparing unaligned tiles is a harder task (with correspondingly lower levels of accuracy) than a comparison of aligned tiles, with the framing given by the context of the other tiles in the same column mitigating some of this difficulty.
@fig-tasks-explain gives an overview of the comparisons of tasks 1 through 3 and the closest corresponding tasks in Cleveland and McGill.
\kr{The updated sketch is great here!! One note -- can we move the label for the "positions along the same, but unaligned scale", it's not aligned with the others}
\hh{that was on purpose :(  that goes beyond Cleveland and McGill ... I'll see what I can do about that}

>>>>

<!-- What is our main objective? -->
<!--
\kr{Should the abstract match -- or be close to -- our SDSS short abstract?}
\hh{I don't think it has to}
-->

What do viewers see when we show them a data chart?
A data chart -- at its core -- maps values to graphical elements: quantitative elements are represented as measurable features, such as position, size, or shade. \hh{do we need to also go into qualitative variables?}
Modern data visualizations are much more than a simple, objective mapping of values to a plane; they contain contextual and design elements, and are often structured to support the viewer in understanding a particular view of a set of data or specific pattern underlying the values.
The design of a data visualization impacts a viewer's ability to achieve that understanding; a poorly designed data visualization may leave viewers struggling to understand the content or context, or make it difficult to complete accurate and useful comparisons of values across groups or time points.
More broadly, the design of a data visualization can change how viewers interact with the chart.

@@

Cleveland and McGill's work -- while a foundational user study in graphical perception -- utilized a small convenience sample, consisting of only a few individuals recruited from among the authors' coworkers and their spouses.
@heerCrowdsourcingGraphicalPerception2010 reproduced Cleveland and McGill's rankings using a larger sample from a crowd sourcing platform, employing a total of 82 Amazon Mechanical Turk (mTurk) workers for the study.
Crowd sourced samples were shown by @borgoCrowdsourcingInformationVisualization2017 to be biased towards more male, younger, and relatively higher education relative to the adult U.S. population as a whole; generally, convenience and opt-in samples are not representative of the general population, a common target audience for data visualization and scientific communication work.
Further, the populations' emphasis on higher education individuals also leads to results which hold for groups of individuals who may be more likely to have prior exposure to data visualization in the context of scientific communication, or more exposure to data topics in higher education, but may not hold across other groups within the population.
<!-- Bias from crowd-sourcing? sample size?  -->

<!--
\kr{The mackinlay thing feels like maybe we are overemphasizing the ranking done by cleveland/mcgill and then mackinlay - I think we could emphasize less if we want}
\hh{my point was to emphasize that Mackinlay is a foundational paper (rightfully so), but not based on any data beyond Cleveland}
-->

@@

<!--Our work centers on studying data visualization design choices and their impact on viewer behavior and accuracy of viewers' responses.-->
Here, we seek to answer whether it is possible to reproduce some of the previous findings in the context of a survey with a large, nationally-representative set of respondents, and within that context we focus on the following research questions:

@@

Specifically, we present viewers with structural variations of bar charts and ask them to answer questions comparing the size of elements within those charts.
In this work, we present the series of tests we completed and the resulting findings.
The remainder of the paper is organized as follows: first, we describe the design of the visual stimuli used in the perception tests.
We then describe the population of study respondents and obtained survey sample.
Subsequently, we share a summary of the resulting survey responses across each of the tests, including analyses on accuracy of responses and response behavior.
Finally, we discuss implications of this work and next steps.

@@

The comparison between sets of marked pairs is intentionally designed to be a difficult task, with the difference between the values represented in the two marked pieces being chosen close to their just-noticeable difference.
The **Just-Noticeable Difference** (JND) is defined as the smallest difference that will be detected 50% of the time.
Prior results from studies on bar charts and pie charts [@luModelingJustNoticeable2022] inform the differences in charts shown to  survey panelists.

@@

```{=tex}
\hh{Not sure where should put this, parking it here for right now:}
\hh{expectations - start}
```
What we call 'aligned' and 'unaligned', here, is similar to Cleveland and McGill's positions along aligned and unaligned scales, but with some modifications: both 'aligned' and 'unaligned' bars share the same axis.
Aligned tiles are additionally anchored in the same position in one dimension, i.e. the difference between their sizes can be reduced to a positional assessment, which in the absence of three-dimensional cues seems to be the general strategy for evaluating barcharts [@zacksReadingBarGraphs1998a].  
Unaligned tiles do not share this anchor, however, the context of the other tiles in the chart provide a frame, which *should* help with an assessment of the tiles' sizes.

We would expect that comparing unaligned tiles is a harder task (with correspondingly lower levels of accuracy) than a comparison of aligned tiles, with the framing given by the context of the other tiles in the same column mitigating some of this difficulty.
@fig-tasks-explain gives an overview of the comparisons of tasks 1 through 3 and the closest corresponding tasks in Cleveland and McGill.
\kr{The updated sketch is great here!! One note -- can we move the label for the "positions along the same, but unaligned scale", it's not aligned with the others}
\hh{What about the sketch now?}

<!--
@cejaTruthSquareAspect2021 suggest that the aspect ratio of areas causes biases in the viewers' assessments -- given that all of the stimuli are close in size (and therefore close in aspect ratio), we do not expect this to be an issue. 
-->

### 2023-07-24 analysis/analysis.qmd
some of these changes are just github's inability to deal with individual sentences

>>>>

From @zacksReadingBarGraphs1998a a good explanation why the results from the facetted barcharts are so bad: we are running into a context-induced illusion. Lines are perceived to be bigger, if there is a bigger line next to them. -- Here, we see the opposite effect.

> The graphical context in which an element occurs can produce distortions in judgments of color, angle, size, and orientation (see, e.g., Goldstein, 1989, especially chap. 7; Howard, 1982). One particularly relevant example of such a distortion is given in Figure 1. This figure demonstrates the parallel lines illusion: When two parallel lines are viewed, the viewer tends to perceive assimilation (the lengths of the lines seem closer than they are) or contrast (the lengths of the lines seem more different than they are), depending on the ratio of the line lengths and the distance between them (Jordan & Schiano, 1986; Schiano, 1986).
